General purpose gene annotation using deep learning

### 2020-01-07 paper/main.tex
paper

\begin{methods}
\section{Methods}
	Helixer models work based on genomic data that was preprocessed and checked for errors by Geenuff \citealp{Denton19}, which stores all information in a SQL database. From there, training and evaluation data is generated. The generated data arrays can be found in Table \ref{Tab:01}. We worked with 192 animal and 60 plant genomes. The data is comprised of all animals genomes in ensembl Metazoa 45 (?) \citealp{Howe19} as well as all non-embargoed plant genomes of the JGI Phytosome 13 database (?, citation). Both groups were treated seperately both during training as well as evaluation. At the start of our work we split both genome groups into training and test sets, which contained 19 or 6 genomes, respectively. Some of the remaining genomes were used for training purposes and all were used for evaluating our models.

>>>>

\begin{methods}
\section{Methods}
We worked with 192 animal and 60 plant genomes. The data is comprised of all animals genomes in ensembl Metazoa 45 (?) \citealp{Howe19} as well as all non-embargoed plant genomes of the JGI Phytosome 13 database (?, citation). Both data groups were used seperately during training as well as during evaluation. At the start of our work we split both genome groups into training and test sets, which contained 19 or 6 genomes, respectively. Some of the remaining genomes were used for training purposes and all were used for evaluating our models.

We preprocessed and stored the genomic information by using Geenuff \citealp{Denton19}. Geenuff is a datachecking and exploration tool, which stores all information inside a SQL database. Training and evaluation data was generated by querying this database and transforming the returned data into a numerical format suitable for machine learning. The generated data arrays can be found in Table \ref{Tab:01}. We only queried the longest transcript of each gene and did not include FASTA sequences that had no functional annotation (explanation why this is a sign of error?). As Geenuff checks the genomic annotations for potential errors like missing UTR sections and marks areas as potentially erroneous, we were able to effectively blank out those bases during training. For the training, we divided each continous genomic sequence into 20.000 or 30.000 long subsequences, which are then used as input or label for the neural network. We appended zero padding if the subsequences are not long enough. If a subsequence is fully marked as erroneous, we do not include it in the training or evaluation datasets.

The selection of training genomes required balancing multiple conflicting tradeoffs. On the one hand, we do want to train with as much data as possible but not all data has a good enough quality to enable a powerful generalization. It is also very desirable to have a broad spread in phylogenetic distance as well as genome size and average gene length as the model is increasingly unlikely to generalize well beyond the borders given by the training data. It was also important to us to get experimental results within a couple of days as we tested many different data and model configurations.
